---
title: "Classification"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
```

# Overview
Add knn, naive bayes, logistic regression, classification trees

# Logistic Regression
# K-Nearest Neighbors (KNN)
KNN works by considering the K training data points that are closest (typically using Euclidean distance) to the test observation.  Then, the test observation is predicted to be in the same class as the majority of the closest training points.  In the event of a tie, one of the classes is randomly chosen.  It is necessary to normalize (generally min-max normalization instead of standardization) the data first since we will be calculating distances.

## KNN: Iris Example
Normalize the iris data and split into a training and test set
```{r}
normalize <- function(x) {
return((x - min(x)) / (max(x) - min(x)))
}
set.seed(123)
irisNormalized = as.data.frame(lapply(iris[,c(1,2,3,4)], normalize))
training_ind = sample(1:nrow(iris), 0.75 * nrow(iris))
train = irisNormalized[training_ind,]
test = irisNormalized[-training_ind,]
train_class = iris[training_ind,5]
test_class = iris[-training_ind,5]
```

Now run knn function
```{r}
library(class)
test_prediction = knn(train, test, cl=train_class, k=3)
```

Calculate accuracy of test data
```{r}
table(test_prediction, test_class)
mean(test_prediction == test_class)*100 #accuracy
```

Try a larger value of k (consider more neighbors)
```{r}
test_prediction = knn(train, test, cl=train_class, k=11)
table(test_prediction, test_class)
mean(test_prediction == test_class)*100 #accuracy
```
In this case, a larger k increased the accuracy of the model, however this was inconsistent across different splits of the training and test sets.  We could use cross-validation to test different values of k (and different train/test sets) to determine the optimal number of neighbors to use.

In general, a smaller value of k means noise in the data will have a higher influence on the boundary causing the boundary to be more flexible.  A larger value of k will have smoother decision boundaries which may not pick up on some boundaries of the data, and is more computationally expensive.

A general rule is to choose $k=\sqrt{N}$ where N is the number of samples in the training data.

# Naive Bayes
Bayes theorem is given as:
$$p(C_k|\textbf{x}) = \frac{p(C_k)p(\textbf{x}|C_k)}{p(\textbf{x})}$$ 
where \textbf{x} represents the features in the data, $C_k$ is a possible class, and $p(C_k|\textbf{x})$ is the probability of belonging to a class given the independent variables defined by \textbf{x}.  This is sometimes rewritten as 
$$\text{posterior} = \frac{\text{prior}\times \text{likelihood}}{\text{evidence}}$$

The denominator of the fraction is constant, and the numeriator is equivalent to the joint probability model $p(C_k,x_1,...,x_n)$.  

If we assume all features in \textbf{x} are mutually independent (this assumption is what makes it *naive*), we can rewrite Bayes theorem as 
$$p(C_k|x_1,...,x_n) \propto p(C_k) \prod_{i=1}^n p(x_i|C_k)$$ 

## Naive Bayes Classifier
Assign a class label $\hat{y} = C_k$ for some k where
$$\hat{y} = \text{argmax}_{k \in \{1,...,K\}}p(C_k)\prod_{i=1}^n p(x_i|C_k)$$

## Naive Bayes: Titanic Example  
Naive Bayes is good to use when all the predictors are categorical, but can also be used on continuous predictors.  It also works well when there are more than 2 classes to predict, unlike other classification methods (e.g. logistic regression).  This example was inspired by [Naive Bayes Blog Post](https://www.r-bloggers.com/understanding-naive-bayes-classifier-using-r/)

We try to predict if a passenger survived on the Titanic given the predictors: Class (1st, 2nd, 3rd, crew), Sex (M,F), and Age (child, adult).
```{r }
data("Titanic")
Titanic = as.data.frame(Titanic)
#Data currently summary data, expand for analysis
repeating_sequence=rep.int(seq_len(nrow(Titanic)), Titanic$Freq)
Titanic_df=Titanic[repeating_sequence,]
Titanic_df = Titanic_df %>% select(-Freq)
```
Fit Naive Bayes model
```{r}
library(e1071)
mod.NB = naiveBayes(Survived~., data=Titanic_df)
mod.NB
```
Make predictions and check accuracy
```{r}
Titanic_df$PredictedSurvived = predict(mod.NB, Titanic_df)
table(Titanic_df$PredictedSurvived, Titanic_df$Survived)
mean(Titanic_df$PredictedSurvived==Titanic_df$Survived)
```
The model predicts if a passenger survives with about 77.8% accuracy.

## Laplace Correction
Suppose there was a set of predictors with a 0 probability of one of the outcomes.  For example, suppose none of the crew survived.  Without any correction, our model will always predict a crew member will not survive, leaving no chance for an unforeseen circumstance.

```{r, warning=F}
missingData = unique(Titanic_df %>% filter(Class=='Crew', Survived=='Yes') %>% select(Class:Survived))
Titanic_df_Less = Titanic_df %>% select(Class:Survived) %>% anti_join(missingData, by = c('Class', 'Sex', 'Age', 'Survived'))
mod.NBNoLaplace = naiveBayes(Survived ~ ., data=Titanic_df_Less, laplace=0)
print(missingData[1,])
predict(mod.NBNoLaplace, missingData[1,], type="raw", threshold=0)
```
 We introduce the Laplace corection, which adds 1 (or whatever defined value) to all the classes.  This forces all joint probabilities to be greater than 0 and allows a (sometimes small) possibility for all outcomes.

```{r, warning=F}
mod.NBLaplace = naiveBayes(Survived ~ ., data=Titanic_df_Less, laplace=1)
predict(mod.NBLaplace, missingData[1,], type="raw", threshold=0)
```








# Classification Trees