---
title: "Regression"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview
Regression typically refers to identifying a (linear) relationship between a set of explanatory variables and response variables.  If the response variable is categorical, this typically refers to an ANOVA model.

## Correlation
Correlation describes the strength and direction of a straight-line (linear) relationship between pair of variables.  The Pearson correlation coefficient is defined as 
$$r = \frac{cov(X,Y)}{s_X s_Y}$$ where $s_X$ is the standard deviation of X, or equivalently, 
$$r = \frac{1}{n-1} \sum z_xz_y$$ where $z_x = \frac{x-\bar{x}}{s_x}$.  We can think of each $z_x$ as the number of standard deviations the data is above the mean.

Correlation ranges from -1 to 1, values with a larger magnitude indicate a stronger correlation, and the sign designates the direction.  If the correlation is 0 we can conclude the variables are not linearly dependent.  However, even if the correlation is small (in magnitude), there may exist a non-linear relationship between them.  Correlation is unit free, unit invariant, and sensitive to outliers.

# Simple Linear Regression
Simple linear regression occurs when there is one independent variable and one dependent variable (typically both continuous).

## Assumptions
Simple linear regression assumes a model of the form:
$$y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$$

We also make various assumptions when fitting a linear regression model:

1. All errors ($\varepsilon_i$'s) are independent
2. Mean of $\varepsilon$ at any fixed x is 0, so average of all $\varepsilon$ is 0
3. At any value of x, the spread of the y's is the same as any other value of x $\rightarrow$ Homoscedasticity
    + $Var(\varepsilon_{ij}) = \sigma^2 = MSE$
4. At any fixed x, the distribution of $\varepsilon$ is normal

We generally assume assumptions 1 and 2 are true, and have methods of verifying assumptions 3 and 4 (explored with examples).

## Fitting the Model
To fit a linear regression model, we want to minimize the sum of squares residuals or sum of squared estimate of errors: $SSE = \sum_{i=1}^n e_i^2 = \sum\limits_{i=1}^n(y_i-\hat{y}_i)^2$ where $\hat{y}_i$ are the fitted values.  Using calculus, it can be shown the solution to this criteria is:

$$\hat{\beta_1} = \frac{\sum\limits_{i=1}^n(X_i-\bar{X})(Y_i-\bar{Y})}{\sum\limits_{i=1}^n(X_i-\bar{X})^2}, \beta_0 = \bar{Y}-\hat{\beta}_1\bar{X}$$

Alternatively, we can use a linear algebra approach and look for a least squares solution to $A\bar{x} = \bar{b}$ where 
$$A = \begin{bmatrix}
1 & x_1\\
1 & x_2\\
\vdots & \vdots\\
1 & x_n
\end{bmatrix}, x = \begin{bmatrix}
\beta_0 \\
\beta_1
\end{bmatrix}, b = \begin{bmatrix}
y_1\\
y_2\\
 \vdots\\
y_n
\end{bmatrix}$$.

However, $\bar{b}$ is not in the column space of $A$, so we must search for a solution for $A\bar{x} = \bar{b}_{||} = \bar{b}-\bar{b}_\perp$.  Multiplying by the transpose of $A$ we have: $A^TA\bar{x} = A^T\bar{b} - A^T\bar{b}_\perp = A^T\bar{b} + 0$.  Finally, solving for $\bar{x}$ we have: $\bar{x} = (A^TA)^{-1}A^T\bar{b}$.


## Simple Linear Regression: Cars Example
We will use the *cars* data to predict the stopping distance of a car given it's speed.  Begin by plotting the data and calculating the correlation:
```{r carplot, message=F, warning=F}
library(ggplot2)
ggplot(cars, aes(speed, dist)) + geom_point()
cor(cars)
```

The relationship appears linear and the correlation is sufficiently large (0.807), so we proceed with fitting a model.

```{r carmodel}
mod.LinearCar = lm(dist~speed, data=cars)
summary(mod.LinearCar)
```

The coefficients, as well as their standard error and `importance` measure are given.  We are also given the *Residual standard error* = $\hat{\sigma}$ and *Multiple R-squared* which in this case (since it's a simple linear model) is equal to the square of the correlation.

### Making Predictions
If we wanted to predicted the *average* stopping distance for a car with speed = 15, we can construct a *confidence* interval with:
```{r car conf}
predict(mod.LinearCar, data.frame(speed=15), se.fit=T, interval='confidence',level=0.95)$fit
```

However, we could also predict the stopping distance for a *particular* car with speed = 15 by constructing a *prediction* interval:
```{r car pred}
predict(mod.LinearCar, data.frame(speed=15), se.fit=T, interval='prediction',level=0.95)$fit
```

The estimates in both cases are the same, however the *confidence* interval only considers variation from repeated experiments while a *prediction* interval considers this variance **and** the variation of an individual.  For either case, the prediction will be more reliable near the center of the data.

#### Adding Line to Plot
We can easily add a linear model to our plot in the ggplot2 package.  This also defaults to drawing a 95% confidence interval.

```{r carplot2, message=F, warning=F}
ggplot(cars, aes(speed, dist)) + geom_point() + geom_smooth(method='lm')
```



### Checking Model Assumptions
#### Homoscedasticity
We assumed the spread of y's is the same at any value of x.  To check this, 
```{r}
plot(fitted.values(mod.LinearCar), residuals(mod.LinearCar), pch=16, xlab='Predicted Value', ylab='Residual')
abline(h=0, lty=2)
```

The residuals should be close to 0 and not have any pattern.  If a pattern does exist (e.g. a funnel/triangle shape), we have evidence of non-homoscedasticity which could potentially be fixed by a data transformation.

#### Normality of Residuals
We also assumed our errors were normally distributed.  To check this we will construct a QQ-plot:
```{r carqq}
qqnorm(residuals(mod.LinearCar), pch=16)
qqline(residuals(mod.LinearCar), col = "red", lwd = 2)
```

In this plot, *Sample Quantiles* refer to the actual value of the residuals, while *Theoretical Quantiles* are the z-scores of the residuals.  If the residuals are normally distributed, the points should fall on/close to the reference line.


# Multiple Linear Regression
Multiple linear regression is an extension of simple linear regression when there are several independent variables or functions of independent variables.  For example, we could have two independent variables ($y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \varepsilon_i$), a quadratic term ($y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i1}^2 + \varepsilon_i$), or including an interaction term ($y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i1}x_{i2} + \varepsilon_i$).  There can be non-linearity in x as long as the coefficients ($\beta_i$) maintain a linear relationship.


## Multiple Linear Regression: ____ Example
model assumptions, test coefficients, multicolinearity

## Assessing Model Fit
leverage, influence, cp, pressp, bic, aic -> just some of these


# Categorical Response (ANOVA model)
## Confidence Intervals
## ANOVA: ____ Example


# Other Related Topics

* Kruskal-Wallis Test
* Dunn Test




Model performance: RMSE, R^2
Leverage: 
library(broom)
mod <- lm(HR ~ SB, data = regulars)
mod %>%
augment() %>%
arrange(desc(.hat)) %>% # or .cooksd
select(HR, SB, .fitted, .resid, .hat) %>%
head()