---
title: "Clustering Overview"
knit: (function(input_file, encoding) {
  out_dir <- '../docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'ClusteringOverview.html'))})
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Finds observations that are similar to one another and groups them together.  Unsupervised technique: don't have defined/known groups.  Must decide to scale data, choose how to calculate distance between observations, and choose how many groups to have.

## Calculating Distance
Consider 3 points on a grid: (1,4), (3,2), (7,8) and calculate the distance between them
```{r}
points = data.frame(x=c(1,3,7), y=c(4,2,8))
dist(points) #defaults to euclidean
dist(points, method="manhattan")
```

### Importance of Scaling
Consider calculating the distance between 3 height (inches) and weight (pounds) pairs.
```{r}
height_weight = data.frame(height = c(65, 68, 60), weight = c(140, 175, 115))
dist(height_weight)
```
Now scale (normalize) the height and weight pairs to be centered at 0 with standard deviation 1 and calculate distance
```{r}
height_weight_scaled = scale(height_weight)
dist(height_weight_scaled)
```
With the unscaled data, observations 1 and 3 are the closest, while with the scaled data observations 1 and 2 are the closest.

### Calculating Distance for Categorical Variables
First convert to dummy variables
```{r message=F, warning=F}
college_gender = data.frame(college = as.factor(c('ASC', 'ASC', 'BUS', 'BUS', 'BUS', 'ENG')), gender = as.factor(c('female','male','male','male','female','male')))
#install.packages('dummies')
library(dummies)
college_gender_dummy = dummy.data.frame(college_gender)
print(college_gender)
print(college_gender_dummy)
```
This is just one example of creating dummy variables in R

Then calculate distance
```{r}
dist(college_gender_dummy, method='binary')
```

## Clustering Techniques
Hierarchical, K-means, PAM, Principal Component Analysis (PCA)